Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-06-14 parallel-tractability-J-v9/background.tex
unify the usage of 'rule instantiation'

We discuss the main issues in this paper using standard datalog notions.
In datalog \cite{database}, a \emph{term} is a variable or a constant. An \emph{atom} $A$
is defined by $A\equiv p(t_1,\ldots,t_n)$ where $p$ is a \emph{predicate} (or \emph{relational})
name, $t_1,\ldots,t_n$ are terms, and $n$ is the arity of $p$. If all the terms in an atom $A$ are
constants, then $A$ is called a \emph{ground atom}.
A datalog \emph{rule} is of the form: `$B_1,\ldots,B_n\rightarrow H$',\footnote{In datalog rules, a comma
represents a Boolean conjunction `$\wedge$'.} where $H$ is referred to as
the \emph{head atom} and $B_1,\ldots,B_n$ the \emph{body atoms}. Each variable in the head atom
of a rule must occur in at least one body atom of the same rule. A \emph{fact} is a rule of
the form `$\rightarrow H$', i.e., a rule with an empty body and the head $H$ being a ground atom.
A datalog program  $P=\langle R, \textbf{I}\rangle$ consists of a
finite set of rules $R$ and a finite set of facts $\textbf{I}$.
A \emph{substitution} $\theta$ is a partial mapping of variables to constants.
For an atom $A$, $A\theta$ is the result of replacing each variable $x$ in $A$
with $\theta(x)$ if the latter is defined. We call $\theta$ a \emph{ground substitution}
if each defined $A\theta$ is a ground atom.
A \emph{ground instantiation} of a rule is obtained by applying a ground substitution on all
the terms in this rule with respect to a finite set of constants occurring in $P$.
Furthermore the ground instantiation of $P$, denoted by $P^*$,
consists of all ground instantiations of rules in $P$.
The predicates occurring only in the body  of some rules are called \emph{EDB predicates},
while the predicates that may occur as head atoms are called \emph{IDB predicates}.

>>>>

We discuss the main issues in this paper using standard datalog notions.
In datalog \cite{database}, a \emph{term} is a variable or a constant. An \emph{atom} $A$
is defined by $A\equiv p(t_1,\ldots,t_n)$ where $p$ is a \emph{predicate} (or \emph{relational})
name, $t_1,\ldots,t_n$ are terms, and $n$ is the arity of $p$. If all the terms in an atom $A$ are
constants, then $A$ is called a \emph{ground atom}.
A datalog \emph{rule} is of the form: `$B_1,\ldots,B_n\rightarrow H$',\footnote{In datalog rules, a comma
represents a Boolean conjunction `$\wedge$'.} where $H$ is referred to as
the \emph{head atom} and $B_1,\ldots,B_n$ the \emph{body atoms}. Each variable in the head atom
of a rule must occur in at least one body atom of the same rule. A \emph{fact} is a rule of
the form `$\rightarrow H$', i.e., a rule with an empty body and the head $H$ being a ground atom.
A datalog program  $P=\langle R, \textbf{I}\rangle$ consists of a
finite set of rules $R$ and a finite set of facts $\textbf{I}$.
A \emph{substitution} $\theta$ is a partial mapping of variables to constants.
For an atom $A$, $A\theta$ is the result of replacing each variable $x$ in $A$
with $\theta(x)$ if the latter is defined. We call $\theta$ a \emph{ground substitution}
if each defined $A\theta$ is a ground atom.
A \emph{ground instantiation} of a rule is obtained by applying a ground substitution on all
the terms in this rule with respect to a finite set of constants occurring in $P$.
For simplicity, we use a \emph{rule instantiation} to denote a ground instantiation of a rule.
Furthermore the ground instantiation of $P$, denoted by $P^*$,
consists of all rule instantiations with respect to $P$.
The predicates occurring only in the body  of some rules are called \emph{EDB predicates},
while the predicates that may occur as head atoms are called \emph{IDB predicates}.

### 2018-06-14 parallel-tractability-J-v9/evaluation.tex
unify the usage of 'rule instantiation'

We implement a prototype system ParallelDHL for DHL$(\circ)$ materialization
based on Algorithm~$\mathsf{A}_{\text{prc}}$. Since ParallelDHL is
evaluated on a
platform which has limited memory space and a fixed number of processors,
the parallel assumption given in Section~\ref{sec:alg-bsc} does not apply to ParallelDHL.
In the implementation of ParallelDHL, we use a hash function to map any rule instance
to the identifier of some processor. Thus, when performing ParallelDHL,
each processor handles a group of rule instances.

@@

We name the above new ontology PTU.
Based on the newly added axioms in PTU, we further modify the ontology-generator
of LUBM such that a \emph{reference chain} of articles can be generated as follows:
$\texttt{referTo}(\texttt{a}_i,\texttt{a}_{i+1})$ and
$\texttt{cite}(\texttt{a}_i,\texttt{a}_{i+1})$ ($i\in\{0,1,2,...\}$),
where $\texttt{a}_i$ denotes an article instance.
As discussed in Section~\ref{sec:lubm-yago},
the original core ontology of LUBM is tractable in parallel via
Skolemization. Further, it can be checked that the concepts in PTU
are all simple concepts. Thus, PTU belongs to the parallel tractability class.
For comparison, we create another core ontology, named NPTU, which
is not tractable in parallel. NPTU is almost the same as PTU except that
Axiom~\eqref{ptu:a2} in PTU is replaced by:
\begin{align}
\exists\texttt{cite}.\texttt{CollegeArticle}\sqsubseteq\texttt{CollegeSession}\label{nptu:a1}\tag{$\beta_4$}
\end{align}
which describes that any article citing a college article is published in a
college session.
It can be checked that NPTU does not satisfy the simple-concept
restriction following analogous arguments to those given in Example~\ref{exp:dhl}.

>>>>

We implement a prototype system ParallelDHL for DHL$(\circ)$ materialization
based on Algorithm~$\mathsf{A}_{\text{prc}}$. Since ParallelDHL is
evaluated on a
platform which has limited memory space and a fixed number of processors,
the parallel assumption given in Section~\ref{sec:alg-bsc} does not apply to ParallelDHL.
In the implementation of ParallelDHL, we use a hash function to map any rule instantiation
to the identifier of some processor. Thus, when performing ParallelDHL,
each processor handles a group of rule instantiations.

@@

We name the above new ontology PTU.
Based on the newly added axioms in PTU, we further modify the ontology-generator
of LUBM such that a \emph{reference chain} of articles can be generated as follows:
$\texttt{referTo}(\texttt{a}_i,\texttt{a}_{i+1})$ and
$\texttt{cite}(\texttt{a}_i,\texttt{a}_{i+1})$ ($i\in\{0,1,2,...\}$),
where $\texttt{a}_i$ denotes an article instantiation.
As discussed in Section~\ref{sec:lubm-yago},
the original core ontology of LUBM is tractable in parallel via
Skolemization. Further, it can be checked that the concepts in PTU
are all simple concepts. Thus, PTU belongs to the parallel tractability class.
For comparison, we create another core ontology, named NPTU, which
is not tractable in parallel. NPTU is almost the same as PTU except that
Axiom~\eqref{ptu:a2} in PTU is replaced by:
\begin{align}
\exists\texttt{cite}.\texttt{CollegeArticle}\sqsubseteq\texttt{CollegeSession}\label{nptu:a1}\tag{$\beta_4$}
\end{align}
which describes that any article citing a college article is published in a
college session.
It can be checked that NPTU does not satisfy the simple-concept
restriction following analogous arguments to those given in Example~\ref{exp:dhl}.

### 2018-06-14 parallel-tractability-J-v9/parallelTractabilityDatalog.tex
unify the usage of 'rule instantiation'

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{\text{bsc}}$) that constructs a materialization graph for a given datalog program.
Our computational model is a PRAM (parallel, random-access machine) that is mostly used to analyze parallel
complexity. This model allows us to introduce \emph{the parallel assumption}:
for any datalog program $P=\langle R, \textbf{I}\rangle$,
any substitution of some atom and any rule instance in $P^*$ can be mapped to a unique memory
location; further, a one-to-one relation can be established between processors and rule instances.
Under this assumption, a processor can check the applicability of its corresponding
rule instance and access the state of an atom occurring in this rule instance in constant time.
Algorithm~$\mathsf{A}_{\text{bsc}}$ is then given as follows.\\

@@

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{\text{bsc}}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any predicate in $\textbf{I}$, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instances in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instances is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{\text{bsc}}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus,
the whole computing time of Algorithm~$\mathsf{A}_{\text{bsc}}$ turns out to be $c_1+\ell\cdot c_2$
where $\ell$ denotes the number of iterations of \ref{alg1:updateG}.

>>>>

In this part, we propose a parallel algorithm
(denoted by Algorithm~$\mathsf{A}_{\text{bsc}}$) that constructs a materialization graph for a given datalog program.
Our computational model is a PRAM (parallel, random-access machine) that is mostly used to analyze parallel
complexity. This model allows us to introduce \emph{the parallel assumption}:
for any datalog program $P=\langle R, \textbf{I}\rangle$,
any substitution of some atom and any rule instantiation in $P^*$ can be mapped to a unique memory
location; further, a one-to-one relation can be established between processors and rule instantiations.
Under this assumption, a processor can check the applicability of its corresponding
rule instantiation and access the state of an atom occurring in this rule instantiation in constant time.
Algorithm~$\mathsf{A}_{\text{bsc}}$ is then given as follows.\\

@@

We now discuss the parallel complexity of Algorithm~$\mathsf{A}_{\text{bsc}}$.
Given a class of datalog programs $\mathbb{P}$ where
a rule set is shared for each datalog program $P=\langle R, \textbf{I}\rangle$ in $\mathbb{P}$.
Let $e$, $v$ and $r$ represent the maximum arity of any predicate in $\textbf{I}$, the maximum number of variables
in any datalog rule, and the number of datalog rules respectively.
We then have that the number of
constants is at most $|\textbf{I}|e$, and the number of all possible rule instantiations in $P^*$
is at most $r(|\textbf{I}|e)^v$.
Note that $e, v$ and $r$ depend only on the rule set $R$ and not on the fact set $\textbf{I}$.
Thus, the memory space for storing the atoms and the rule instantiations is polynomial in the size of $\textbf{I}$.
This also means that the number of processors is polynomially bounded.
The computing time of \ref{alg1:addFacts} and \ref{alg1:halt} occupy constant time (denoted by $c_1$) because of parallelism.
Since Algorithm~$\mathsf{A}_{\text{bsc}}$ works under the parallel assumption, one iteration of
\ref{alg1:updateG} also costs constant time (denoted by $c_2$). Thus,
the whole computing time of Algorithm~$\mathsf{A}_{\text{bsc}}$ turns out to be $c_1+\ell\cdot c_2$
where $\ell$ denotes the number of iterations of \ref{alg1:updateG}.
